A Recurrence

### 2014-11-27 draft.tex
added last section

\newtheorem{proposition}{Proposition}

@@

In Section 2 we explore a simplification of the question; we only consider $F$ of the form $F(x,y)=f(y)/x$. If we assume that $f$ is a polynomial we can completely classify such $F$ of period 2,3,4,5 and 6. We also show that for any $n>0$ we can produce a $f(y)$ such that $F$ has period $n$.  In Section 3 we look at another simplification of the question; we consider $F$ of the form $f(x,y)/g(x,y)$ where $f$ and $g$ are linear polynomials. We are able to completely describe $F$ of period 2 and 3.

@@

@@

It is somewhat obvious that $F(x,y)=x$ is the only ratio of linear functions that has period 2, however for larger $n$ the question and answer are slightly more interesting. 
\begin{proposition} 
If $F(x,y)$ is of the form in Equation \ref{rationallinear} and the recurrence relation it produces has period $3$ then $F$ is one of three following functions:
\begin{align*}
F(x,y) &= -x-y \\
F(x,y) &= \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) y \\
F(x,y) &= \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) y.
\end{align*}
\end{proposition}
\begin{proof}
In order for $F(x,y)$ to induce a recurrence relation with period $3$ we must have $a_3=a_0$. So using the expression in Equation \ref{athree} we must have 
\begin{align*}
\left( (c_1 x + c_2 y +c_3) c_1 y + c_2 (b_1 x + b_2 y + b_3) + c_3(c_1 x + c_2 y +c_3) \right) x = \\(c_1 x + c_2 y + c_3) b_1 y + b_2(b_1 x + b_2 y + b_3) + b_3 (c_1 x = c_2 y + c_3),
\end{align*}
If we multiply this expression out we get 
\begin{align*}
c_1^2 x^2 y + c_2 c_1 y^2 x + (c_3 c_1 + c_2 b_2 + c_3 c_2)xy + (c_2b_1 + c_3 c_1) x^2 + (b_3 + c_3)x = \\ c_1 b_1 xy + c_2 b_1 y^2 + c_3 b_1 y + (b_2 b_1 + b_3 c_1)x + (b_2^2 +b_3 c_2)y + b_2 b_3 + b_3 c_3.
\end{align*}
Because this relation must hold for arbitrary $x$ and $y$ we know that terms with the same exponents for $x$ and $y$ must cancel. So for instance we must have 
\begin{equation*}
c_1^2 x^2 y = 0, 
\end{equation*}
and so we must have $c_1=0$. We can do this for terms of each order and find the following 5 relations for the coefficients 
\begin{align}
\label{fiverelations}
c_2(b_2 + c_3) =0 \\
c_2 b_1 = 0 \\
b_3 + c_3 = b_2 b_1 \\
c_3 b_1 + b_2^2 + b_3 c_2 =0 \\
b_3(b_2 + c_3) = 0. 
\end{align}
Suppose that $c_2 =0$, then $F$ is of the form 
\begin{equation*}
F(x,y) = \frac{ b_1 x + b_2 y + b_3}{c_3}.
\end{equation*}
Therefore we must have 
\begin{equation*}
c_3^2 x = b_1 c_3 y + b_2 (b_1 x + b_2 y + b_3) + b_3,
\end{equation*}
which gives us a new family of relations 
\begin{align*}
b_2 b_1 = c_3^2 \\
c_3 b_1 + b_2^2 =0.
\end{align*}
Combining these two gives us 
\begin{equation*}
c_3^3 + b_2^3 = 0, 
\end{equation*}
which has three solutions $c_3 = -b_2$ and $c_3 = \frac{1}{2} ( b_2 \pm i\sqrt{3}b_2)$. Note that $c_3 = -b_2$ implies that we must have $b_1=b_2$ which gives us $F(x,y) = -x -y$. It is easy to check that this is in fact periodic with period 3. On the other hand $c_3 = \frac{1}{2} ( b_2 \pm i\sqrt{3}b_2)$ implies that $b_1 = \frac{1}{2} ( -b_2 \pm i\sqrt{3}b_2)$, which gives us 
\begin{align*}
F(x,y) &= \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) y \\
F(x,y) &= \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) y.
\end{align*}
It is also easy to check that these two $F$ have period 3. This exhausts the possibilities when $c_2=0$ so let us asssume otherwise. Then based on the relations \ref{fiverelations} we must have 
\begin{equation*}
b_1 = 0,
\end{equation*}
and so $F(x,y)$ takes the form 
\begin{equation*}
F(x,y) = \frac{b_2 y + b_3}{c_2 y + c_3}.
\end{equation*}
Therefore if $a_3 = a_0$ we must have that 
\begin{equation*}
x = \frac{ b_2^2 y + b_2b_3 +b_3(c_2 y + c_3)}{c_2 b_2 y + c_2 b_3 + c_3 c_2 y + c_3^2},
\end{equation*}
but the left hand side depends only on $x$ and the right hand side depends only on $y$. This relation cannot hold for arbitrary $x,y$ and so we cannot have such $F$. Therefore we must have $c_2=0$ and so the only $F$ with period 3 are the ones we have given. 
\end{proof}
\end{document}

>>>>

\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}

@@

In Section 2 we explore a simplification of the question; we only consider $F$ of the form $F(x,y)=f(y)/x$. If we assume that $f$ is a polynomial we can completely classify such $F$ of period 2,3,4,5 and 6. In Section 3 we look at another simplification of the question; we consider $F$ of the form $f(x,y)/g(x,y)$ where $f$ and $g$ are linear polynomials. We are able to completely describe $F$ of period 2 and 3. In the final section, we show a method that given any period, (we conjecture) will produce a function of the form $F(x,y) = y^n/x$ which attains that period. For this last part, we do not assume that $n$ is an integer.

@@

It is somewhat obvious that $F(x,y)=x$ is the only ratio of linear functions that has period 2, however for larger $n$ the question and answer are slightly more interesting. 
\begin{proposition} 
If $F(x,y)$ is of the form in Equation \ref{rationallinear} and the recurrence relation it produces has period $3$ then $F$ is one of three following functions:
\begin{align*}
F(x,y) &= -x-y \\
F(x,y) &= \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) y \\
F(x,y) &= \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) y.
\end{align*}
\end{proposition}
\begin{proof}
In order for $F(x,y)$ to induce a recurrence relation with period $3$ we must have $a_3=a_0$. So using the expression in Equation \ref{athree} we must have 
\begin{align*}
\left( (c_1 x + c_2 y +c_3) c_1 y + c_2 (b_1 x + b_2 y + b_3) + c_3(c_1 x + c_2 y +c_3) \right) x = \\(c_1 x + c_2 y + c_3) b_1 y + b_2(b_1 x + b_2 y + b_3) + b_3 (c_1 x = c_2 y + c_3),
\end{align*}
If we multiply this expression out we get 
\begin{align*}
c_1^2 x^2 y + c_2 c_1 y^2 x + (c_3 c_1 + c_2 b_2 + c_3 c_2)xy + (c_2b_1 + c_3 c_1) x^2 + (b_3 + c_3)x = \\ c_1 b_1 xy + c_2 b_1 y^2 + c_3 b_1 y + (b_2 b_1 + b_3 c_1)x + (b_2^2 +b_3 c_2)y + b_2 b_3 + b_3 c_3.
\end{align*}
Because this relation must hold for arbitrary $x$ and $y$ we know that terms with the same exponents for $x$ and $y$ must cancel. So for instance we must have 
\begin{equation*}
c_1^2 x^2 y = 0, 
\end{equation*}
and so we must have $c_1=0$. We can do this for terms of each order and find the following 5 relations for the coefficients 
\begin{align}
\label{fiverelations}
c_2(b_2 + c_3) =0 \\
c_2 b_1 = 0 \\
b_3 + c_3 = b_2 b_1 \\
c_3 b_1 + b_2^2 + b_3 c_2 =0 \\
b_3(b_2 + c_3) = 0. 
\end{align}
Suppose that $c_2 =0$, then $F$ is of the form 
\begin{equation*}
F(x,y) = \frac{ b_1 x + b_2 y + b_3}{c_3}.
\end{equation*}
Therefore we must have 
\begin{equation*}
c_3^2 x = b_1 c_3 y + b_2 (b_1 x + b_2 y + b_3) + b_3,
\end{equation*}
which gives us a new family of relations 
\begin{align*}
b_2 b_1 = c_3^2 \\
c_3 b_1 + b_2^2 =0.
\end{align*}
Combining these two gives us 
\begin{equation*}
c_3^3 + b_2^3 = 0, 
\end{equation*}
which has three solutions $c_3 = -b_2$ and $c_3 = \frac{1}{2} ( b_2 \pm i\sqrt{3}b_2)$. Note that $c_3 = -b_2$ implies that we must have $b_1=b_2$ which gives us $F(x,y) = -x -y$. It is easy to check that this is in fact periodic with period 3. On the other hand $c_3 = \frac{1}{2} ( b_2 \pm i\sqrt{3}b_2)$ implies that $b_1 = \frac{1}{2} ( -b_2 \pm i\sqrt{3}b_2)$, which gives us 
\begin{align*}
F(x,y) &= \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) y \\
F(x,y) &= \left( \frac{1}{2} +\frac{i\sqrt{3}}{2} \right) x + \left( \frac{1}{2} -\frac{i\sqrt{3}}{2} \right) y.
\end{align*}
It is also easy to check that these two $F$ have period 3. This exhausts the possibilities when $c_2=0$ so let us asssume otherwise. Then based on the relations \ref{fiverelations} we must have 
\begin{equation*}
b_1 = 0,
\end{equation*}
and so $F(x,y)$ takes the form 
\begin{equation*}
F(x,y) = \frac{b_2 y + b_3}{c_2 y + c_3}.
\end{equation*}
Therefore if $a_3 = a_0$ we must have that 
\begin{equation*}
x = \frac{ b_2^2 y + b_2b_3 +b_3(c_2 y + c_3)}{c_2 b_2 y + c_2 b_3 + c_3 c_2 y + c_3^2},
\end{equation*}
but the left hand side depends only on $x$ and the right hand side depends only on $y$. This relation cannot hold for arbitrary $x,y$ and so we cannot have such $F$. Therefore we must have $c_2=0$ and so the only $F$ with period 3 are the ones we have given. 
\end{proof}

\section{The $f(y) = y^n$ Case}

\emph{Written by Fermi Ma and Perry Kleinhenz, edited by Erik Waingarten}\\

Consider the case where $F(x,y) = f(y)/x$ where $f(y) = y^n$. Then the sequence can be written as follows
\[ x^1y^0, x^0y^1, x^{-1}y^{n}, x^{-n}y^{n^2-1}, x^{-n^2+1}y^{n^3-2n},x^{-n^3+n^2+n}y^{n^4-3n^2+1}, \dots \]

The interesting thing to note is that all of the terms in the sequence are now monomials. This makes the analysis much easier. Now let $\alpha_k$ be the sequence of the exponents of $x$ in the above sequence, and let $\beta_k$ be the sequence of exponents of $y$. So we have
\begin{align*}
\alpha_0 &= 1  &\beta_0& = 0\\
\alpha_1 &= 0 &\beta_1& = 1\\
\alpha_2 &= -1  &\beta_2& = n\\
\alpha_3 &= -n &\beta_3& = n^2-1\\
\alpha_4 &= -n^2+1 &\beta_4& = n^3-2n\\
\alpha_5 &= -n^3+2n &\beta_5& =n^4-3n^2+1\\ 
\alpha_6 &= -n^4+3n^2 -1 &\beta_6& =n^5-4n^3+3n \\ 
\alpha_7 &= -n^5+4n^3 - 3n &\beta_7& =n^6-5n^4+6n^2 - 1
\end{align*}

We can use these $\{\alpha_i\}$ and $\{\beta_i\}$ sequences to get values of $n$ that achieve certain periods. For example, if we want a function of the form $F(x,y) = y^n/x$ with period 5, we simply set $\alpha_5 = \alpha_0 = 1$, $\beta_5 = \beta_0 = 0$, $\alpha_6 = \alpha_1 = 0$, and $\beta_6 = \beta_1 = 1$. This gives us a system of four equations in one variable
\begin{align*}
-n^3+2n &= 1\\
n^4-3n^2+1&=0\\
-n^4+3n^2 -1&= 0\\
n^5-4n^3+3&n=1\\
\end{align*}
There is no guarantee that this system has a solution, but it turns out that setting $n = -\frac{1+\sqrt{5}}{2} = \phi$, the golden ratio, works. This tells us that $F(x,y) = y^{-\phi}/x$ is periodic with period 5, and we can indeed check that this is the case.

In our numerical calculations, it seems like this system of equations always \emph{does} has a real solution, which leads us to conjecture that for any period strictly greater than 2, there is a function of the form $y^n/x$ with that period.

\subsection{Properties of $\alpha_k$ and $\beta_k$}

We first notice that $\alpha_k$ and $\beta_k$ are similar sequences, as they can be defined as
\begin{align*}
\alpha_0 = 1, \alpha_1 = 0, \alpha_k = n\alpha_{k-1} - \alpha_{k-2}\\
\beta_0 = 0, \beta_1 = 1, \beta_k = n\beta_{k-1} - \beta_{k-2}
\end{align*}

The recurrence relation follows from simply inspecting the function $F(x,y) = y^n/x$.

Note also that $-\beta_k = \alpha_{k+1}$. This follows from the fact that this holds true from looking at the $k =0$ and $k = 1$ cases, and then noticing that the recurrence relations preserve this property.

We also note that $\beta_k$ can be written as 
\begin{align*}
\binom{k-1}{k-1} n^{k-1} - \binom{k-2}{k-3}n^{k-3} +\binom{k-3}{k-5}n^{k-5} -\binom{k-3}{k-7}n^{k-7} + \dots \\
=\sum_{i = 0}^{\lfloor (k-1)/2 \rfloor} (-1)^i\binom{k-1-i}{k-1-2i}n^{k-1-2i}\\
\end{align*}
for $k \geq 1$.

To check this, note that it holds for $k = 1$ and $k = 2$, and then plug it in to the recurrence relation
\begin{align*}
\beta_k =& n\beta_{k-1} - \beta_{k-2}\\
\sum_{i = 0}^{\lfloor (k-1)/2 \rfloor} (-1)^i\binom{k-1-i}{k-1-2i}n^{k-1-2i} = & n \sum_{i = 0}^{\lfloor (k-2)/2 \rfloor} (-1)^i\binom{k-2-i}{k-2-2i}n^{k-2-2i} \\
& +\sum_{i = 0}^{\lfloor (k-3)/2 \rfloor} (-1)^i\binom{k-3-i}{k-3-2i}n^{k-3-2i}\\
\end{align*}
Checking this equation for when $k$ is even and when $k$ is odd, as well as using combinatorial identities, shows that this is true.

So we have an expression for $\beta_k$, and this in turn gives us an expression for $\alpha_k$ as noted above. Thus, we know that if we want to attain a specific period $p$, we have to solve equations involving these known expressions.

It remains to show that these equations always produce a real solution for $n$.

@@

\end{document}
